Data organization in spreadsheets

### 2017-02-24 manuscript.Rmd
Revisions in response to reviewer 2

In this paper we offer practical recommendations for organizing data in
spreadsheets. By following this advice, researchers will create spreadsheets
that are less error-prone, easier for computers to process, and easier to share
with collaborators and the public. Spreadsheets that adhere to our
recommendations will interface easily with the tidy tools and reproducible
methods described elsewhere in this collection and will form the basis of a
robust and reproducible analytic workflow.

@@

But note that Microsoft Excel does
bizarre things with
dates (see <https://storify.com/kara_woo/excel-date-system-fiasco>). It
stores them internally as an integer, counting the days since
1900-01-01. Wait that is only for Windows; on Macs, it counts the days
since 1904-01-01.

So, be careful to check that the dates haven't been mangled when your
data come out of Excel.

@@

```{r 07_two_header_rows, fig.width=10, fig.cap="A spreadsheet with two header rows. It is better to have a single header row (see Figure 8)."}
mat6 <- rbind(c("", "", "week 4", "", "", "week 6", "", "", "week 8", "", ""),
              c("Mouse ID", "SEX", rep(c("date", "weight", "glucose"), 3)),
              c("3005", "M", "3/30/2007",  19.3, 635.0, "4/11/2007",  31,   460.7, "4/27/2007",  39.6, 530.2),
              c("3017", "M", "10/6/2006",  25.9, 202.4, "10/19/2006", 45.1, 384.7, "11/3/2006",  57.2, 458.7),
              c("3434", "F", "11/22/2006", 26.6, 238.9, "12/6/2006",  45.9, 378.0, "12/22/2006", 56.2, 409.8),
              c("3449", "M", "1/5/2007",   27.5, 121.0, "1/19/2007",  42.9, 191.3, "2/2/2007",   56.7, 182.5),
              c("3499", "F", "1/5/2007",   19.8, 220.2, "1/19/2007",  36.6, 556.9, "2/2/2007",   43.6, 446.0))

@@

Have sympathy for your analyst (which could be yourself):
organize your data as a rectangle (or, if necessary, as a set of
rectangles).

@@

Consider using a formal version control system, like
git, though it is not the best for data
files. If you want to get a bit fancy, maybe look at
dat (<https://https://datproject.org/>).

>>>>

Spreadsheets are often used as a multipurpose tool for data entry, storage,
analysis, and visualization. Most spreadsheet programs allow users to perform
all of these tasks, however we believe that spreadsheets are best suited to data
entry and storage, and that analysis and visualization should happen separately.
Analyzing and visualizing data in a separate program reduces the risk of
contaminating or destroying the raw data in the spreadsheet.

In this paper we offer practical recommendations for organizing spreadsheet data
in a way that both humans and computer programs can read. By following this
advice, researchers will create spreadsheets that are less error-prone, easier
for computers to process, and easier to share with collaborators and the public.
Spreadsheets that adhere to our recommendations will interface easily with the
tidy tools and reproducible methods described elsewhere in this collection and
will form the basis of a robust and reproducible analytic workflow.

@@

But note that Microsoft Excel does bizarre things with dates
(see <https://storify.com/kara_woo/excel-date-system-fiasco>). It stores them
internally as an integer, counting the days since 1900-01-01. Wait, that is only
for Windows; on Macs, it counts the days since 1904-01-01. So, be careful to
check that the dates haven't been mangled when your data come out of Excel.

@@

```{r 07_two_header_rows, fig.width=10, fig.cap="A spreadsheet with two header rows. It is better to have a single header row. See Figure 8 for a tidy data layout that eliminates the need for multiple header rows and repeated column headers."}
mat6 <- rbind(c("", "", "week 4", "", "", "week 6", "", "", "week 8", "", ""),
              c("Mouse ID", "SEX", rep(c("date", "weight", "glucose"), 3)),
              c("3005", "M", "3/30/2007",  19.3, 635.0, "4/11/2007",  31,   460.7, "4/27/2007",  39.6, 530.2),
              c("3017", "M", "10/6/2006",  25.9, 202.4, "10/19/2006", 45.1, 384.7, "11/3/2006",  57.2, 458.7),
              c("3434", "F", "11/22/2006", 26.6, 238.9, "12/6/2006",  45.9, 378.0, "12/22/2006", 56.2, 409.8),
              c("3449", "M", "1/5/2007",   27.5, 121.0, "1/19/2007",  42.9, 191.3, "2/2/2007",   56.7, 182.5),
              c("3499", "F", "1/5/2007",   19.8, 220.2, "1/19/2007",  36.6, 556.9, "2/2/2007",   43.6, 446.0))

@@

Have sympathy for your analyst (which could be yourself): organize your data as
a rectangle (or, if necessary, as a set of rectangles). This raises the question
of whether to use multiple worksheets within one Excel file, or whether it would
be better to use multiple files, each with one worksheet. We prefer to have
multiple files with one sheet each so we can save the data as CSV files (more on
this later), but if you do use multiple worksheets in a file be sure to use a
consistent structure.

@@

Consider using a formal version control system, like
git, though it is not the best for data
files. If you want to get a bit fancy, maybe look at
dat (<https://datproject.org/>).
